% Created 2015-09-14 Mon 13:04
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\author{Giulio Bottazzi}
\date{\today}
\title{Betting and Optimal Portfolios}
\hypersetup{
  pdfkeywords={},
  pdfsubject={},
  pdfcreator={Emacs 24.3.1 (Org mode 8.2.4)}}
\begin{document}

\maketitle
\tableofcontents


\section{Introduction}
\label{sec-1}

In his '52 seminal paper Kelly proposes an original connection between
information theory and betting strategies. His work revealed the
possibility to investigate the problem of forecasting from a very
general perspective. One of the topics of the following research has
been to investigate the possibility to devise a strategy that
outperforms a given set of competing strategies under minimal
information about the nature and dynamics of the underline process.
\section{Assignment}
\label{sec-2}

Your assignment is divided in two tasks. Each task starts from a
specific paper which proposes a way of dealing with the problem of
building, under certain conditions and in a specific sense, the
"optimal" strategy or family of strategies. You are required to
understand and implement, using a new database, the idea discussed in
the paper. A comparative exercise of the proposed family of strategies
with other alternatives will be highly appreciated.

\subsection{Task 1: Betting}
\label{sec-2-1}

In 1974 Cover proposes a \href{https://drive.google.com/file/d/0B3G1Di9qZKBmVGhtendHaWtheWM/view?usp=sharing}{Universal Gambling Scheme} that should, to a
certain extent, perform the magic of beating any single strategy
competitors propose. From \href{http://www.football-data.co.uk}{football-data} website you can download the
time series of matches with the final outcome and odds proposed by
different online gambling sites. You can see the odds proposed by the
different bookmakers as forecasts on the outcome of the
game. Investigate and compare their performance and, starting from the
advice of these "experts", try to build a strategy, or "betting
scheme", that leads to better asymptotic performances. Hint: Italian
\emph{Serie A} championship is consistently available from 2002-03.

\subsection{Task2 : Investing}
\label{sec-2-2}

Not satisfied by having dominated the betting arena, in 1991 Cover
moves to conquer the financial market with his \href{https://drive.google.com/file/d/0B3G1Di9qZKBmMFRHMkRmZGJ4R00/view?usp=sharing}{Universal
Portfolios}. From \href{http://finance.yahoo.com/}{yahoo!finance} download daily and weekly time series
of a portfolio of stocks. Compare the performance of different
buy-and-hold and re-balancing strategies, possibly using out-of-sample
analysis. Then try to implement Cover suggested strategy and discuss
its performance. Hint: Pick a set of liquid, high capitalization, and
diverse assets. A possible choice could be: Bank of America
Corporation (BAC), General Electric Company (GE), Ford Motor Co. (F),
Citigroup, Inc. (C), Pfizer Inc. (PFE), Hewlett-Packard Company (HPQ),
JPMorgan Chase \& Co. (JPM), AT\&T, Inc. (T), Wells Fargo \& Company
(WFC) and J. C. Penney Company, Inc. (JCP).

\section{Bibliography}
\label{sec-3}

\subsection{Essential}
\label{sec-3-1}

\begin{enumerate}
\item Thomas M. Cover (1974) "Universal Gambling Schemes and the
Complexity Measures of Kolmogorov and Chaitin", Technical report
n.12, Stanford University.

\item Thomas M. Cover (1991) "Universal Portfolios", Mathematical Finance
1 (1): 1–29.
\end{enumerate}
\subsection{Further reading}
\label{sec-3-2}

\begin{enumerate}
\item J. L. Kelly , Jr. (1956) \href{https://drive.google.com/file/d/0B3G1Di9qZKBmdUpUZDNDOG83eHM/view?usp=sharing}{"A New Interpretation of Information Rate"},
The Bell System Technical Journal.

\item Leo Breiman (1961) \href{https://drive.google.com/file/d/0B3G1Di9qZKBmbXotb21Od08wdjA/view?usp=sharing}{"Optimal Gambling Systems for Favorable Games"},
Proc. Fourth Berkeley Symp. on Math. Statist. and Prob.

\item Meir Feder (1991) \href{https://drive.google.com/file/d/0B3G1Di9qZKBmNUVKQm5pU0ZSUmM/view?usp=sharing}{"Gambling using a finite state machine"}, IEEE
Transactions in Information Theory.
\end{enumerate}
% Emacs 24.3.1 (Org mode 8.2.4)
\end{document}